\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-29 syntax.tex
Adding intro for substitution calculus

We formalize our language as a deep embedding in a type theory with QIIT types, every judgement is a (dependent) QIIT \textit{type} and all judgement derivations are well-typed QIIT terms. (Otherwise how can they be encoded in Agda?) There are four standard judgement forms in MLTT,
well-typed contexts $\goodCtx{\Gamma}{l}$, well-typed types
$\goodType{\Gamma}{T}{l}$, well-typed terms $\goodTerm{\Gamma}{t}{T}$, and
well-typed substitutions $\goodSub{\Gamma}{\gamma}{\Delta}$. These four judgements (QIIT types) are
mutually recursively defined together. Our presentation
follows the style of~\citet{kaposi2019gluing}, but for readability we
omit universe levels attached to judgements.
%\YZ{What's the distinction of judgments vs. sequents?}.\EDJreply{I am not sure about the correct terminology here. Here I want to say "even underivable judgement has to be well-typed". That's the main reason I spell out the fake Agda code there. I actually want the reader to have intrinsic-typed syntax in mind all the time (like read conventional math but acknowledge we are manipulating Agda)}\EDJreply{I rephrase that so there won't be confusion for the future readers. Please check}
Latex-wise, we spell out these \textit{well-formedness rules} of type
judgements and their Agda counterparts to make things more accessible. Of
course, a well-formed judgement is not necessarily derivable, just like
a well-typed QIIT type is not necessarily inhabited.

@@

We quickly review the substitution calculus,

@@

Now we show selected typing rules: we will have dependent function types
and dependent pair types. Dependent pair type is used to model Coq
module, which is useful when we need to model the ``compilation'' from
linkages to modules. We also have $\beta,\eta$-rule for both dependent
function type and dependent pair type.
The substitution law for these two
require \textit{de bruijn lifting}: for
$\goodTerm{\Gamma}{\sigma}{\Delta}$, we define its de Bruijn lifting
$\goodTerm{\Gamma,A[\sigma]}{\sigma^\uparrow}{\Delta, A}$ given
$\goodType{\Delta}{A}{}$.\YZ{What is A[sigma] and what is de Bruijn lifting? Seems like if the reader is not familiar with the substitution calculus, then they will get lost starting here. Broadly speaking, a gentle intro to the intrinsic setting would help the reader a great deal.}

@@

$$
\Rule[name=Base Type/Term Subst]
{\goodSub{\Gamma}{\gamma}{\Delta}}
{\goodType{\Gamma}{(El \ T)[\gamma] \equiv El \ (T[\gamma]) }{j} \quad
 \goodTerm{\Gamma}{"tt"[\gamma] \equiv "tt"}{\cB} \quad 
 \goodTerm{\Gamma}{"ff"[\gamma] \equiv "ff"}{\cB} 
}
$$
\judgebox{\goodSub{\Gamma}{\sigma}{\Delta}}
$$
\Rule[name=Empty Subst]
{}{\goodSub{\Gamma}{\epsilon}{\cdot}}
\quad
\Rule[name=Id Subst]
{}{\goodSub{\Gamma}{"id"}{\Gamma}}
\quad
\Rule[name=Subst Extension]
{\goodSub{\Gamma}{\sigma}{\Delta} \quad \goodTerm{\Gamma}{t}{A[\sigma]}}
{\goodSub{\Gamma}{(\sigma, t)}{(\Delta, A)}}
$$

>>>>

We formalize our language as a deep embedding in a type theory with QIIT types. Every judgement is a (dependent) QIIT \textit{type} and all judgement derivations are well-typed QIIT terms. (Otherwise how can they be encoded in Agda?) There are four standard judgement forms in MLTT,
well-typed contexts $\goodCtx{\Gamma}{l}$, well-typed types
$\goodType{\Gamma}{T}{l}$, well-typed terms $\goodTerm{\Gamma}{t}{T}$, and
well-typed substitutions $\goodSub{\Gamma}{\gamma}{\Delta}$. These four judgements (QIIT types) are
mutually recursively defined together. Our presentation
follows the style of~\citet{kaposi2019gluing}, but for readability we
omit universe levels attached to judgements.
%\YZ{What's the distinction of judgments vs. sequents?}.\EDJreply{I am not sure about the correct terminology here. Here I want to say "even underivable judgement has to be well-typed". That's the main reason I spell out the fake Agda code there. I actually want the reader to have intrinsic-typed syntax in mind all the time (like read conventional math but acknowledge we are manipulating Agda)}\EDJreply{I rephrase that so there won't be confusion for the future readers. Please check}
Latex-wise, we spell out these \textit{well-formedness rules} of type
judgements and their Agda counterparts to make things more accessible. Of
course, a well-formed judgement is not necessarily derivable, just like
a well-typed QIIT type is not necessarily inhabited.

@@

We quickly review the substitution calculus. Substitution is very (if not the most) important because function is first-class citizen, and can be applied (almost) everywhere and return (almost) everything, thus we require the resulting substitution to commutate with most (if not all) type/term constructs.

A term $\goodTerm{\Gamma}{t}{A}$ can be understood as one term $t$ of type $A$, with a bunch of holes (of type $\Gamma$) inside the term $t$. A substitution $\goodSub{\Gamma}{\gamma}{\Delta}$ can be similarly understood in the way that $\gamma$ will output a \textbf{series} of (dependently-typed related) terms of types $\Delta$, parametrized by $\Gamma$ (i.e. we can consider there are holes in $\gamma$ of type $\Gamma$). Thus (1) \ruleref{Sub Extend} looks like concatenating one series of terms with a new term (the weird looking type of $\goodTerm{\Gamma}{t}{A[\sigma]}$ is because we need $A$ to appear in the context of $\Delta$ so that we can use \ruleref{Context Extension} to create new context); (2) we have \ruleref{Proj Subst-1} to extract a subpart of a given substitution; (3) \ruleref{Proj-Ext} explains why substitution really looks like a (deeply-nested) pair.

Since the $\Gamma$ before the $\vdash$ is always indicating the types of ``holes'', this explains the \ruleref{Term Subst} and \ruleref{Type Subst}: given a term $\goodTerm{\Delta}{t}{A}$ and a substitution $\goodSub{\Gamma}{\gamma}{\Delta}$, we will have a new term $\goodTerm{\Gamma}{t[\gamma]}{A[\gamma]}$ after the substitution---this substitution can be understood as ``filling all the holes''. Once all the holes are filled with a given substitution, the new holes all come from the subsitution itself. Thus consecutive subsitution can be composed as \ruleref{Sub Comp}, because all the holes are always from the last substitutions.

Note that, for simplicity, we will (a lot of time) writing $\pi_1$ as a \textbf{shorthand} of $\goodSub{\Gamma,A}{\pi_1~"id"}{\Gamma}$, and similarly another shorthand for $\goodTerm{\Gamma, A}{\pi_2 := \pi_2~"id"}{A[\pi_1]}$.

@@

We will have dependent function types
and dependent pair types. Dependent pair type is used to model Coq
module, which is useful when we need to model the ``compilation'' from
linkages to modules. We also have $\beta,\eta$-rule for both dependent
function type and dependent pair type.
The substitution law for these two
require \textit{de bruijn lifting}: for a given substitution
$\goodTerm{\Gamma}{\sigma}{\Delta}$ and a type
$\goodType{\Delta}{A}{}$, we define its de Bruijn lifting
$\goodTerm{\Gamma,A[\sigma]}{\sigma^\uparrow := (\sigma \circ \pi_1, \pi_2)}{\Delta, A}$ . Intuitively speaking, de Bruijn lifting is to ``fill all the holes except the last one''.\YZ{What is A[sigma] and what is de Bruijn lifting? Seems like if the reader is not familiar with the substitution calculus, then they will get lost starting here. Broadly speaking, a gentle intro to the intrinsic setting would help the reader a great deal.}\EDJreply{A quick intuitive explanation of substitution-calculus is given above. Please check. }

@@

$$
\Rule[name=Base Type/Term Subst]
{\goodSub{\Gamma}{\gamma}{\Delta}}
{\goodType{\Gamma}{(El \ T)[\gamma] \equiv El \ (T[\gamma]) }{j} \quad
 \goodTerm{\Gamma}{"tt"[\gamma] \equiv "tt"}{\cB} \quad 
 \goodTerm{\Gamma}{"ff"[\gamma] \equiv "ff"}{\cB} 
}
$$
\judgebox{\goodSub{\Gamma}{\sigma}{\Delta}}
$$
\Rule[name=Ept Subst]
{}{\goodSub{\Gamma}{\epsilon}{\cdot}}
\quad
\Rule[]
{}{\goodSub{\Gamma}{"id"}{\Gamma}}
\quad
\Rule[name=Sub Comp]{
  \goodSub{\Delta}{\delta}{\Theta}
  \quad \goodSub{\Gamma}{\gamma}{\Delta} 
}{\goodSub{\Gamma}{\delta \circ \gamma}{\Theta}}
\Rule[name=Sub Extend]
{\goodSub{\Gamma}{\sigma}{\Delta} \quad \goodTerm{\Gamma}{t}{A[\sigma]}}
{\goodSub{\Gamma}{(\sigma, t)}{(\Delta, A)}}
$$
